### 2014-10-17 paper/include/background.tex
flikker started

The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}. This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval. This approach is called \textit{burst refresh} and blocks normal accesses during the burst time, where power consumption peaks. A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.

@@

As refresh operations are made by row, the refresh operations will increase linearly with the device capacity. In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}. Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}. When the device capacity reaches $64\:GB$, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

>>>>

\todo[inline]{change the next paragraph to introduce retention time.}
The memory controller (MC) normally issues refresh operations periodically at an interval of \textit{64~ms} for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}. This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the \textit{64~ms} interval. This approach is called \textit{burst refresh} and blocks normal accesses during the burst time, where power consumption peaks. A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.

@@

As refresh operations are made by row, the refresh operations will increase linearly with the device capacity. In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}. Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}. When the device capacity reaches \textit{64~GB}, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

### 2014-10-17 paper/include/refrint.tex
flikker started

This technique is proposed by Agrawal et al. \cite{refrint} and is based on Access Recency. It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well. The concept is based on two types of unnecessary refreshes which originate from so called cold and hot rows. Cold rows are those used far apart in time or not used at all. The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated. Hot rows consist of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.

@@

Four different policies can be applied for the cold rows, where the policies decide what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}. \textit{All} refreshes every row, regardless of whether it is valid or not. \textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row. \textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated. $n$ and $m$ are time-out counters that are decremented upon refresh.

\todo[inline]{Drafty text below!}

Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows. Visibility correspond to the amount of cache row accesses made by the application. If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used. A high visibility and a small footprint works best with small $n$ and $m$, whereas a large footprint is likely more usefull with small \textit{n} and \textit{m}.

The technique modifies the MC. Which modifications and how large overhead? 5 bit snapshot for each row + state and n+m counters. some logic as well.

>>>>

This technique is proposed by Agrawal et al. \cite{refrint} and is based on Access Recency and Validity. It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented in DRAM as well. The concept is based on two types of unnecessary refreshes which originate from so called cold and hot rows. Cold rows are those used far apart in time or not used at all. The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated. Hot rows consist of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.

@@

Four different policies can be applied for the cold rows, where the policies decide what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}. \textit{All} refreshes every row, regardless of whether it is valid or not. \textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row. \textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated. Moreover, \textit{WB} is implemented using so called \textit{Count} bits, which is set either to $n$ or $m$ and is decremented upon refresh. % Mention the State bits as well?

Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows, where visibility correspond to the number of cache row accesses made by the application. If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used. A high visibility and a small footprint works best with large $n$ and $m$, whereas a high visibility and a large footprint likely benefits more from small \textit{n} and \textit{m}.

\todo[inline]{Drafty below. Ask Sally, confusing paper.}

Refrint modifies the MC and also stores data on chip(?). In an unmodified MC, there already exist a counter which can be used as the global counter and the authors have chosen to store the $N$ snapshot bits in the MC as well as the rows' valid bits in the MC. From experiments, the authors found out that one or two are good values for $N$. The logic to process the policies for the cold rows consists of decision logic, as well as Count bits and State bits for each cache row. This logic is located on chip(?)/in DRAM(?)(???) and has apparently a negligible overhead. In simulations, $N$ was two bits and Count was 5 bits, giving a total overhead of $0.6\%$, when cache row size is 64 bytes.

### 2014-10-17 paper/include/smart_refresh.tex
flikker started

Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them. The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue. The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of $16\:KB$.

The technique was simulated on $2$ GB of main memory. Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.

>>>>

Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them. The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue. The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of \textit{16~KB}.

The technique was simulated on \textit{2~GB} of main memory. Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.
